```{r setup, include=FALSE, echo=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2, width = 55, digits = 4
)
```

# Preface {.unnumbered}

## Software information and conventions {.unnumbered}

I used the **knitr**\index{knitr} package [@xie2015] and the **quarto**\index{quarto} package [@quarto] to compile my book. My R session information is shown below:

```{r}
xfun::session_info()
```

Package names are in bold text (e.g., **rmarkdown**), and inline code and filenames are formatted in a typewriter font (e.g., `knitr::knit('foo.Rmd')`). Function names are followed by parentheses (e.g., `bookdown::render_book()`).

## About the Author

Alex Gold leads the Solutions Engineering team at [RStudio](www.rstudio.com).

He works with customers of RStudio's professional software to help them deploy, configure, and use RStudio's professional software and open source tooling in R and Python.

In his free time, he enjoys landscaping, handstands, and Tai Chi.

He occasionally blogs about data, management, and leadership at [alexkgold.space](http://alexkgold.space).

## Acknowledgments {.unnumbered}

A lot of people are helping me write this book.

This book is published to the web using GitHub Actions from [rOpenSci](https://github.com/orchid00/actions_sandbox).

## Color palette

Tea Green: CAFFDO Steel Blue: 3E7CB1 Kombu Green: 273c2c Bright Maroon: B33951 Sandy Brown: FCAA67
